---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# r2r

<!-- badges: start -->
[![R-CMD-check](https://github.com/vgherard/r2r/workflows/R-CMD-check/badge.svg)](https://github.com/vgherard/r2r/actions)
<!-- badges: end -->

`r2r` provides a flexible implementation of hash tables in R, allowing for:

- arbitrary R objects as keys and values,
- arbitrary key comparison and hash functions,
- customizable behaviour (throw or return a default values) on missing key exceptions.

## Installation

You can install the development version of `r2r` from [GitHub](https://github.com/vgherard/r2r) with:

``` r
# install.packages("devtools")
devtools::install_github("vgherard/r2r")
```
## Examples

```{r}
library(r2r)
```

#### Basic Manipulations

We create an empty hash map with:

```{r example}
m <- hashmap()
```

We can insert key-value pairs in `m` in several different ways:

```{r}
m[["key"]] <- "value"
m[c(1, 2, 3)] <- c("a", "b", "c") # Vectorized over keys and values
m[[c(4, 5, 6)]] <- c("d", "e", "f") # Not vectorized
```

The following queries explain the differences between the `[[` and `[` operator mentioned in the comments above:

```{r}
m[["key"]]

m[c(1, 2, 3)]
m[[c(1, 2, 3)]]

m[c(4, 5, 6)]
m[[c(4, 5, 6)]]
```

Single element insertions and queries can also be performed through the generics `insert()` and `query()`

```{r}
insert(m, "user", "vgherard") # Modifies `m` in place
query(m, "user")
```

#### Sets

In addition to hash maps, we can also create hash sets, which simply store keys:

```{r}
s <- hashset()
insert(s, 1)
s[[2]] <- T # equivalent to insert(s, 2)
s[c(1, 2, 3)]
```

#### Key and value types

There is no restriction on the type of object you can use as keys and values. For instance:

```{r}
m[[ lm(wt ~ mpg, mtcars) ]] <- list("This is my fit!", 840)
m[[ lm(wt ~ mpg, mtcars) ]]
m[[ lm(cyl ~ mpg, mtcars) ]]
```

#### Setting default values
You can set default values for missing keys. For instance:

```{r}
m <- hashmap(default = 0)
```

which is useful for creating a counter:

```{r}
objects <- list(1, 1, "1", FALSE, "1", 1)
for (object in objects)
	m[[object]] <- m[[object]] + 1
m[["1"]]
```

Alternatively, you may throw an exception upon querying a missing key:

```{r}
m <- hashmap(on_missing_key = "throw")
tryCatch(m[["Missing key"]], error = function(cnd) "Oops!")
```


#### Using custom key comparison and hash functions
`hashmap`s and `hashmap`s use by default `base::identical()` to compare keys. 
For instance:

```{r}
m <- hashmap()
m[[1]] <- "double"
m[["1"]] <- "character"
m[[1]]
```

This behavior can be changed by explicitly providing a key comparison function.
For this to work correctly, one must also explicitly provide an hash function which produces the same hashes for equivalent keys. A simple way to do this is 
to apply a preprocessing function to keys, as illustrated by the following example.

We assume that keys are length one complex numbers, and consider two keys equivalent when they have the same direction in the complex plane. The direction of a complex vector can be found applying the R function `Arg()`, which is thus a sensible key preprocessing function. We can instruct an hashmap to preprocess its keys in this way through the constructor's `key_preproc_fn` argument:

```{r}
m <- hashmap(key_preproc_fn = Arg)
```

Let us check that everything works as intended:

```{r}
m[list(1, 1 + 1i, 1i)] <- list("EAST", "NORTH-EAST", "NORTH")
m[[10]]
m[[100i]]
m[[2 + 2i]]
```
