---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# r2r

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/r2r)](https://CRAN.R-project.org/package=r2r)
[![R-CMD-check](https://github.com/vgherard/r2r/workflows/R-CMD-check/badge.svg)](https://github.com/vgherard/r2r/actions)
[![Codecov test coverage](https://codecov.io/gh/vgherard/r2r/branch/master/graph/badge.svg)](https://codecov.io/gh/vgherard/r2r?branch=master)
<!-- badges: end -->

[`r2r`](https://vgherard.github.io/r2r/) provides a flexible implementation of hash tables in R, allowing for:

- arbitrary R objects as keys and values,
- arbitrary key comparison and hash functions,
- customizable behaviour (throw or return a default value) on missing key exceptions.

## Installation

You can install the development version of `r2r` from [GitHub](https://github.com/vgherard/r2r) with:

``` r
# install.packages("devtools")
devtools::install_github("vgherard/r2r")
```
## Usage

```{r}
library(r2r)
m <- hashmap()

# Insert and query a single key-value pair
m[[ "user" ]] <- "vgherard"
m[[ "user" ]]

# Insert and query multiple key-value pairs
m[ c(1, 2, 3) ] <- c("one", "two", "three")
m[ c(1, 3) ]

# Keys and values can be arbitrary R objects
m[[ lm(mpg ~ wt, mtcars) ]] <- c(TRUE, FALSE, TRUE)
m[[ lm(mpg ~ wt, mtcars) ]]
```

## Getting help

For further details, including an introductory vignette illustrating the features of `r2r` hash maps, you can consult the `r2r` [website](https://vgherard.github.io/r2r/). If you encounter a bug, want to suggest a feature or need further help, you can [open a GitHub issue](https://github.com/vgherard/r2r/issues).

## Comparison with `hash`

CRAN package [`{hash}`](https://CRAN.R-project.org/package=hash) also offers an implementation of hash tables based on R environments. The two tables below offer a comparison between `{r2r}` and `{hash}` (for more details, see the [benchmarks](https://vgherard.github.io/r2r/articles/benchmarks.html) Vignette)

```{r echo=FALSE}
knitr::kable(
	data.frame(
		Feature = c(
			"Basic data structure",
			"Arbitrary type keys", 
			"Arbitrary type values",
			"Arbitrary hash function",
			"Arbitrary key comparison function",
			"Throw or return default on missing keys",
			"Hash table inversion"
		),
		r2r = c("R environment", "X", "X", "X", "X", "X", ""),
		hash = c("R environment", "", "X", "", "", "", "X")
		),
	align = "c",
	caption = "Features supported by {r2r} and {hash}."
	)
```

```{r echo=FALSE}
knitr::kable(
	data.frame(
		Task = c(
			"Key insertion",
			"Key query",
			"Key deletion"
			),
		Comparison = c(
			"{r2r} ~ {hash}",
			"{r2r} < {hash}",
			"{r2r} << {hash}"
			)
		),
	align = "c",
	caption = "Performances of {r2r} and {hash} for basic hash table operations."
	)
```
